Case study validations of automatic bullet matching

### 2018-11-19 index.Rmd
small edits

## Identification Using 3D Imaging Technology
More explicit characterization of bullet surfaces has been discussed since at least 1958 (\citeauthor{davis1968introduction}), but at the time technology was not sufficiently advanced to make an analysis based on 3D measurements or surface traces a viable option. \citet{ComputerIdentificationBullets1978} demonstrated use of a scanning electron microscope to quantitatively examine and compare bullet striae. More recently, approaches using 3D measurement data were explicitly described in 1999 (\citeauthor{dekinderAutomatedComparisonsBullet1999}), and have been further developed by \citet{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of land engraved areas (LEAs), rather than using visual or image comparison techniques. In many cases, these \svp{approaches} also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. Utilizing the 3D measurements allows for examination \svp{of both the peaks and valleys in LEAs; visual examination of valleys is more difficult} except in relation to the peaks.

Commonly, approaches derived from 3D surface measurements use some features which are similar to those visually assessed by FTEs \citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient for an exclusion. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements. One of the most common features used to assess the strength of a match based on individual characteristics is the cross-correlation function, which is \svp{utilized} in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae (CMS) \citep{chuAutomaticIdentificationBullet2013}.

@@

\begin{itemize}
\item[(a)] Identify an area on the LEA with expressed striae. Locate a stable crosscut (see \citet{aoas}) in the identified area.
\item[(b)] Discard extraneous/contaminated data, such as data from groove engraved area and areas affected by break-off or contact with objects after the bullet exited the barrel  (``tank rash").
\item[(c)] Remove bullet curvature \hh{using a Gaussian filter \cite{chuAutomaticIdentificationBullet2013}\svp{This mentions use of a gaussian filter, but not for the specific purpose of removing curvature. Also, since we're discussing the aoas paper, might be better to stick to the non-parametric smooth?} or non-parametric smooth \citep{aoas}}. 
\end{itemize}
A signature for a land engraved area is then defined as the sequence $S(x)$, $x = 1, ..., I$, where $I$ is the number of observed locations across the base of the bullet. \autoref{fig:signatures} shows a set of six signatures corresponding to the six land engraved areas of bullets 1 and 2 from barrel 1 of set 44 of the Hamby study \citep{hamby}.

@@

\svp{Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength.} \hh{Maximized cross-correlation is used to pair-wise align signatures \citet{aoas,vorburgerApplicationsCrosscorrelationFunctions2011}. }

The cross-correlation function (ccf) between two signatures $S_1(x)$ and $S_2(x)$ is defined as $$CC(S_1, S_2, z) = \text{cor}\left(S_1(x), S_2(x + z)\right)$$ where \hh{$x$ and $x+z$ are integer values appropriately defined within the domains of $S_1$ and $S_2$ and }  $z$ is the lag between the two sequences. The lag $z$ used to achieve maximum $CC$ is used to determine the best alignment; the maximal value of $CC$ is  used to assess the match strength between the two sequences. When signatures contain missing values, pairwise complete observations are used to calculate the cross-correlation.

@@

One of the measures firearms examiners use to describe the strength of a match quantitatively is the number of consecutive matching striae (CMS) as defined by \citet{biasotti}.
If there are at least 6 consecutive matching striae, the bullets are considered to be a good match \citep{biasotti_firearms_1997,nichols_scientific_2006}. 
CMS in this setting are counted as the number of consecutive peaks in signatures of two lands that match\svp{, in part because it is difficult to visually assess valleys except as relative to peaks}. 
\hh{Note that because signatures are moved into phase based on the maximum cross-correlation rather than matching striation marks,} \svp{it is possible for two lands to }\hh{have a consecutive matching striae of 0. } \svp{CMS are generally considered to be a very conservative criteria, in that bullets fired through the same barrel may not have a high number of consecutively matching striae; in this sense, CMS may be more likely to produce an inconclusive result.}

\hh{XXX for CMS we need to define peaks, and matching peaks, and for that we need to define the additional filter/smooth.}\svp{XXX Can't we just refer to the aoas paper? It seems to be defined in there relatively well.}

\hh{The random forest presented in \citet{aoas} is based on a combination of multiple characteristics, such as maximized cross-correlation, number of matching striae, and number of consecutively matching striae among others. }

@@

## Identification Errors
When evaluating algorithm performance, it is useful to distinguish between two types of errors, which we will refer to as \textbf{wrong identification} and \textbf{missed identification} in this paper. Wrong identifications are those in which two bullets from different sources are determined to be from the same source. Missed identifications are those in which two bullets from the same source are determined to be from different sources.
\hh{Note that this definition of error is more strict than the AFTE Theory of Identification, in that it does not allow for inconclusive results, but forces a decision based on a cutoff value. }
\svp{XXX Relate to AFTE Theory of Examination criteria? How to handle ``inconclusive"?}
\svp{XXX May want to have the error type table/chart as a visual representation? XXX}

\svp{In this paper, we validate the algorithm described in \citet{aoas} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross-correlation and consecutive matching striae. }

@@

\svp{Random forests \citep{breiman} have} \hh{a built-in internal testing mechanism to prevent potential overfitting. Error rates reported by random forest algorithms are based on these internal test sets. However, there is some discussion on whether or in which situations these error rates are biased \citep{rfOver2, rfOver}. 
Neither of these papers addresses another issue with internal test sets: internal test sets are constructed to have the same distribution as the training data (apart from sampling variability).} \svp{For any machine learning method, a true benchmark of the performance of an algorithm requires testing its performance on external test data.} \hh{Good performance (in terms of low error rates) on external test data validates a ML algorithm. External data also allows an assessment of the algorithm's sensitivity to distributional changes} \svp{as well as testing the algorithm's robustness by going outside the parameters of the training data.  }

@@

\hh{For the validation of the automatic matching algorithm we are considering three validation sets of -- what should be -- increasing difficulty level:} 
\begin{itemize}
\item {\bf Hamby set 44} is one set of the Hamby study \citep{hamby}. Each Hamby set  consists of a total of 35 bullets fired through ten consecutively manufactured barrels of Rugers P85. Each set consists of 20 known bullets (two from each of the ten barrels) and 15 questioned bullets of unknown origin. Note that all Hamby sets are closed sets; that is, all questioned bullets are fired through one of the ten barrels. The ammunition used for this set were 9 mm Luger 115 Grain Full Metal Jacket  from the Winchester Ammunition Company.
\item {\bf Phoenix PD}
Tylor Klep from Phoenix PD provided sets of known test fires and questioned bullets: the set of known bullets consists of three test fires (B1, B2, B3) from each of eight different, consecutively manufactured Ruger P-95 barrels (A9, C8, F6, L5, M2, P7, R3, U10). Ten questioned bullets were provided (B, E, H, J, K N, Q, T, Y, Z). This set is an open set; that is,\ it is not known in advance whether all (or any) of the questioned bullets are fired from the known barrels. \hh{In fact, the results will show that three of the questioned bullets were fired from three different barrels not included in the knowns. These three barrels correspond to an eleventh Ruger P-95, a Ruger P-95C and a Ruger P-85.} \svp{All bullets fired for this study are American Eagle 9mm  copper jacketed Luger.}
Land engraved areas for each of the six lands of each bullet were scanned by Bill Henderson (Sensofar).
\item {\bf Houston FSI}
This study was set up by Melissa Nally and Kasi Kirksey from FSI Houston. Three test sets based on ten consecutively rifled Ruger LCP barrels (A, B, C, D, E, F, G, H, I, J) and three other, non-consecutively rifled Ruger LCP barrels (R1, R2, R3). Each test set consists of three test fires each from five consecutively rifled barrels. Additionally, ten questioned bullets are provided for each kit. The ammunition used in both test fires and the questioned bullets were Remington UMC 9mm Luger Full Metal Jackets. All three of the test sets are open; that is,\ not every one of the questioned bullets is fired from the five known barrels in each of the test set.
The structure of these three sets is similar to a forthcoming study by Nally and Kirksey, but the results here come from preliminary test sets made available to us.
\end{itemize}

>>>>

## Identification Using 3D Imaging Technology
More explicit characterization of bullet surfaces has been discussed since at least 1958 (\citeauthor{davis1968introduction}), but at the time technology was not sufficiently advanced to make an analysis based on 3D measurements or surface traces a viable option. \citet{ComputerIdentificationBullets1978} demonstrated use of a scanning electron microscope to quantitatively examine and compare bullet striae. More recently, approaches using 3D measurement data were explicitly described in 1999 (\citeauthor{dekinderAutomatedComparisonsBullet1999}), and have been further developed by \citet{bachrachDevelopment3DbasedAutomated2002,xieAutomatedBulletidentificationSystem2009, chuPilotStudyAutomated2010}. These approaches utilize 3D surface measurements directly to characterize the topology of land engraved areas (LEAs), rather than using visual or image comparison techniques. In many cases, these approaches also provide some level of automation of the comparison process, with the goal of reducing human biases by augmenting the visual information with 3D measurements. Utilizing the 3D measurements allows for examination of both peaks and valleys in LEAs, \hh{it also allows us to take the depth of striae into account, something that is difficult, if not impossible, for visual examination.}

Commonly, approaches derived from 3D surface measurements use some features which are similar to those visually assessed by FTEs \citep{luAutomatedBulletIdentification2014}. Class characteristics, which are shared by a group of firearms with the same rifling design, manufacturer, and tooling process, are typically evaluated first, as a mismatch on class charcteristics is sufficient for an exclusion. Automated approaches to estimation of width of the land engraved area and twist angle were described in \citet{chuPilotStudyAutomated2010}. Individual characteristics, which are not shared by all members of a class, can also be automatically assessed from 3D surface measurements. One of the most common features used to assess the strength of a match based on individual characteristics is the cross-correlation function, which is utilized in several studies \citep{maNISTBulletSignature2004, vorburgerApplicationsCrosscorrelationFunctions2011, chuPilotStudyAutomated2010}. Additional features proposed for automatically assessing match strength also include signature distance \citep{maNISTBulletSignature2004}, striae depth and width \citep{ComputerIdentificationBullets1978}, and consecutive matching striae (CMS) \citep{chuAutomaticIdentificationBullet2013}.

@@

\begin{itemize}
\item[(a)] Identify an area on the LEA with expressed striae. Locate a stable crosscut (see \citet{aoas}) in the identified area.
\item[(b)] Discard extraneous/contaminated data, such as data from groove engraved area and areas affected by break-off or contact with objects after the bullet exited the barrel  (``tank rash").
\item[(c)] Remove bullet curvature using a non-parametric smooth. 
\end{itemize}
A signature for a land engraved area is then defined as the sequence $S(x)$, $x = 1, ..., I$, where $I$ is the number of observed locations across the base of the bullet. \autoref{fig:signatures} shows a set of six signatures corresponding to the six land engraved areas of bullets 1 and 2 from barrel 1 of set 44 of the Hamby study \citep{hamby}.

@@

Once signatures have been extracted from stable regions with expressed striae, they must be aligned in order to assess their match strength. Maximized cross-correlation is used to pair-wise align signatures \citet{aoas,vorburgerApplicationsCrosscorrelationFunctions2011}.

The cross-correlation function (ccf) between two signatures $S_1(x)$ and $S_2(x)$ is defined as $$CC(S_1, S_2, z) = \text{cor}\left(S_1(x), S_2(x + z)\right)$$ where \hh{$x$ and $x+z$ are integer values appropriately defined within the domains of $S_1$ and $S_2$ and  $z$ is the lag between the two sequences. The lag $z$ used to achieve maximum $CC$ is used to determine the best alignment; the maximal value of $CC$ is  used to assess the match strength between the two sequences. When signatures contain missing values, pairwise complete observations are used to calculate the cross-correlation.

@@

One of the measures firearms examiners use to describe the strength of a match quantitatively is the number of consecutive matching striae (CMS) as defined by \citet{biasotti}.
If there are at least 6 consecutive matching striae, the bullets are considered to be a good match \citep{biasotti_firearms_1997,nichols_scientific_2006}. 
CMS in this setting are counted as the number of consecutive peaks in signatures of two lands that match, in part because it is difficult to visually assess valleys except as relative to peaks. 
Note that because signatures are moved into phase based on the maximum cross-correlation rather than matching striation marks, it is possible for two lands to have a consecutive matching striae of 0. CMS are generally considered to be a very conservative criteria, in that bullets fired through the same barrel may not have a high number of consecutively matching striae; in this sense, CMS may be more likely to produce an inconclusive result.

<!--\hh{XXX for CMS we need to define peaks, and matching peaks, and for that we need to define the additional filter/smooth.}\svp{XXX Can't we just refer to the aoas paper? It seems to be defined in there relatively well.}-->

The random forest presented in \citet{aoas} is based on a combination of multiple characteristics, such as maximized cross-correlation, number of matching striae, and number of consecutively matching striae among others.

@@

## Identification Errors
When evaluating algorithm performance, it is useful to distinguish between two types of errors, which we will refer to as \textbf{wrong identification} and \textbf{missed identification} in this paper. Wrong identifications are those in which two bullets from different sources are determined to be from the same source. Missed identifications are those in which two bullets from the same source are determined to be from different sources.
\hh{Note that this definition of error is more stringent than the AFTE Theory of Identification, in that it does not allow for inconclusive results, but forces a decision based on a cutoff value. }
\svp{XXX Relate to AFTE Theory of Examination criteria? How to handle ``inconclusive"?}
\svp{XXX May want to have the error type table/chart as a visual representation? XXX}

In this paper, we validate the algorithm described in \citet{aoas} on three external test sets. We assess the results on each set and evaluate whether the minimal requirements described above are fulfilled for bullet-to-bullet SAM scores. We compare the random forest algorithm's performance to the performance of other suggested measures for quantitative assessment of match strength, such as cross-correlation and consecutive matching striae.

@@

Random forests \citep{breiman} have a built-in internal testing mechanism to prevent potential overfitting. Error rates reported by random forest algorithms are based on these internal test sets. However, there is some discussion on whether or in which situations these error rates are biased \citep{rfOver2, rfOver}. 
Neither of these papers addresses another issue with internal test sets: internal test sets are constructed to have the same distribution as the training data (apart from sampling variability). For any machine learning method, a true benchmark of the performance of an algorithm requires testing its performance on external test data.Good performance (in terms of low error rates) on external test data validates a ML algorithm. External data also allows an assessment of the algorithm's sensitivity to distributional changes as well as testing the algorithm's robustness by going outside the parameters of the training data.

@@

For the validation of the automatic matching algorithm we are considering three validation sets of -- what should be -- increasing difficulty level: 
\begin{itemize}
\item {\bf Hamby set 44} is one set of the Hamby study \citep{hamby}. Each Hamby set  consists of a total of 35 bullets fired through ten consecutively manufactured barrels of Rugers P85. Each set consists of 20 known bullets (two from each of the ten barrels) and 15 questioned bullets of unknown origin. Note that all Hamby sets are closed sets; that is, all questioned bullets are fired through one of the ten barrels. The ammunition used for this set were 9 mm Luger 115 Grain Full Metal Jacket  from the Winchester Ammunition Company.
\item {\bf Phoenix PD}
Tylor Klep from Phoenix PD provided sets of known test fires and questioned bullets: the set of known bullets consists of three test fires (B1, B2, B3) from each of eight different, consecutively manufactured Ruger P-95 barrels (A9, C8, F6, L5, M2, P7, R3, U10). Ten questioned bullets were provided (B, E, H, J, K N, Q, T, Y, Z). This set is an open set; that is,\ it is not known in advance whether all (or any) of the questioned bullets are fired from the known barrels. \hh{In fact, the results will show that three of the questioned bullets were fired from three different barrels not included in the knowns. These three barrels correspond to an eleventh Ruger P-95, a Ruger P-95C and a Ruger P-85.} \svp{All bullets fired for this study are American Eagle 9mm  copper jacketed Luger.}
Land engraved areas for each of the six lands of each bullet were scanned by Bill Henderson (Sensofar).
\item {\bf Houston FSI}
This study was set up by Melissa Nally and Kasi Kirksey from FSI Houston. Three test sets based on ten consecutively rifled Ruger LCP barrels (A, B, C, D, E, F, G, H, I, J) and three other, non-consecutively rifled Ruger LCP barrels (R1, R2, R3). Each test set consists of three test fires each from five consecutively rifled barrels. Additionally, ten questioned bullets are provided for each kit. The ammunition used in both test fires and the questioned bullets were Remington UMC 9mm Luger Full Metal Jackets. All three of the test sets are open; that is,\ not every one of the questioned bullets is fired from the five known barrels in each of the test set.
The structure of these three sets is similar to a forthcoming study by Nally and Kirksey, but the results here come from preliminary test sets made available to us.
\end{itemize}
